---
author: "Carla Flore"
title: "SVM Classification on Spambase Dataset"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r load-packages, message=FALSE, warning=FALSE}
# ---- Load packages ----
library(caret)
library(doParallel)
```

Below, we load the dataset and ensure that the response variable `Class` is encoded as a binary factor.

```{r load-data}
# ---- Set seed and load data ----
set.seed(4500393)
df <- read.csv("spambase.csv")
df$Class <- as.factor(df$Class)  # Ensure binary factor
```

We split the dataset into 70% training and 30% testing data to evaluate generalization performance.

```{r train-test-split}
# ---- Train-test split (70/30) ----
idx <- sample(1:nrow(df), size = 0.7 * nrow(df), replace = FALSE)
train_raw <- df[idx, ]
test_raw  <- df[-idx, ]
```

SVMs are sensitive to feature scales. We standardize predictors using the training set's mean and variance.

```{r scale-predictors}
# ---- Scale predictors ----
x_train <- scale(train_raw[, -ncol(train_raw)])
x_test  <- scale(test_raw[, -ncol(test_raw)],
                 center = attr(x_train, "scaled:center"),
                 scale = attr(x_train, "scaled:scale"))
```

We reattach the response variable to the scaled predictors for training and testing.

```{r combine-data}
# ---- Combine scaled predictors with target ----
train_svm <- as.data.frame(x_train)
train_svm$Class <- train_raw$Class

test_svm <- as.data.frame(x_test)
test_svm$Class <- test_raw$Class
```

We set up parallel processing to accelerate model tuning.

```{r parallel-setup, message=FALSE}
# ---- Setup parallel backend ----
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

We define the tuning grid for cost and gamma (sigma).

```{r define-grid}
# ---- Define tuning grid ----
tune_grid <- expand.grid(
  C = c(0.1, 1, 10),
  sigma = c(0.01, 0.1, 1)
)
```

The best hyperparameters are selected using 10-fold CV and parallel computation.

```{r tune-model-parallel}
# ---- Cross-validation tuning with caret + parallel ----
set.seed(4500393)
svm_tuned <- train(
  Class ~ .,
  data = train_svm,
  method = "svmRadial",
  trControl = trainControl(method = "cv", number = 10),
  tuneGrid = tune_grid,
  preProcess = NULL
)
```

We stop the parallel backend.

```{r stop-cluster}
# ---- Stop cluster ----
stopCluster(cl)
registerDoSEQ()
```

We apply the trained model to the test set and compute performance metrics.

```{r predict-evaluate-caret}
# ---- Predict and evaluate on test set ----
svm_pred <- predict(svm_tuned, newdata = test_svm)
conf_matrix <- confusionMatrix(svm_pred, test_svm$Class)
```

Finally, we print the best parameters and the confusion matrix.

```{r results}
# ---- Print results ----
print(svm_tuned$bestTune)
print(conf_matrix)
```

The SVM classifier with a radial basis kernel, tuned using 10-fold cross-validation, achieved its best performance with `C = 10` and `sigma = 0.01`. The model showed excellent accuracy on the test set, achieving 93.6% accuracy. It also exhibited high sensitivity (94.7%) and specificity (91.9%), indicating balanced and robust performance in detecting both spam and non-spam emails. The Kappa value of 0.866 suggests strong agreement beyond chance, and the McNemar’s test showed no significant asymmetry in misclassification types. 
```{r summary-table, echo=FALSE}
# ---- Summary table of confusion matrix ----
summary_table <- data.frame(
  Metric = c("Accuracy", "95% CI (lower)", "95% CI (upper)", "Kappa", "Sensitivity",
             "Specificity", "Pos Pred Value", "Neg Pred Value", "Balanced Accuracy"),
  Value = c(0.9363, 0.9221, 0.9486, 0.866, 0.9467, 0.9199, 0.9489, 0.9165, 0.9333)
)

knitr::kable(summary_table, digits = 4, caption = "Summary of SVM Model Performance on Test Set")
```

```{r plot-tuning-results, fig.height=4.5, fig.width=6}
# ---- Visualize tuning results ----
plot(svm_tuned)
```
