---
title: "Spam_group_project"
author: "Bos Noah"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r}
df = read.csv('spambase.csv')

```

# summary statistics
- mean, sd min, max, na
- No NA
- 


```{r}
sapply(df, function(x) c(mean = mean(x, na.rm=TRUE),
                         sd = sd(x, na.rm=TRUE),
                         min = min(x, na.rm=TRUE),
                         max = max(x, na.rm=TRUE),
                         na = sum(is.na(x))))
```

```{r, fig.width=10, fig.high = 10}
cor_matrix = cor(df)
heatmap(cor_matrix, scale = 'column')
```
- Dendograms at the top and right show hieracrical lustering. Maybe 2/3 big clusteres can be seen. After that not clear pattern
- Check the CLASS variable (our outcome). It is moderately correlated with some of the variables, and some not.

# Would PCA be relevant
```{r}
library(EFAtools)
N_FACTORS(df, criteria = c('PARALLEL', 'SCREE', 'EKC', 'KGC'), method = 'ML', eigen_type_other = 'PCA' )
```
- 11 to 15 factors are suitable. Which is quite large 


# distribution of predictors

- boxplots show quite some skewness
- most of the emails contain words 0 times, some emails contain a lot of them. 
```{r}
boxplot(df[,1:(ncol(df)-1)])
```

```{r, fig.width=15, fig.height=10}
library(ggplot2)
library(reshape2)
df_long <- melt(df)
ggplot(df_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal()
```

- more clearly shown here, in most emails 0 words are used in other emails the words get used more often. 
- in text data like emails, the majority of words appear infrequently.
- So: most values are near zero, and a few emails have very high frequencies of certain words.
- Certain words (e.g. “free”, “money”) appear very often in spam and rarely in normal emails, exaggerating skewness in those features.
```{r}
library(dplyr)
library('paran')
df_subset = df |>
  select(-Class)

paran(df_subset, iterations = 5000, graph = T)
```
# CHECKING PCA WITH 15 FACTORS
- not used yet
```{r}
library(psych)
pca_rotated = principal(df_subset, nfactors = 15, rotate = "varimax", scores = TRUE)
print(pca_rotated, digits = 2, cut = 0.3)
```

